NumPy---Array Programming in Python

### 2020-01-31 body.tex
Shape the story of the proliferation section, according to my and Jarrod's discussion

NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It runs on machines ranging from embedded devices to the world's
largest supercomputers, with performance
approaching that of compiled languages.
For algorithms that do not fit the array programming approach,
or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
that transparently accelerate bottlenecks.
These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure Python.
% smooth experience/intengration from high level <-> low level?

@@

NumPy suits most user's needs, but it cannot support every specialized need
of the community; due to its in-memory data model, it is limited in its ability
to work with very large datasets, datasets split across multiple systems, and
computations that require graphics processing units (GPUs), tensor processing
units (TPUs) or other specialized hardware for performance reasons.
However, both distributed data and the parallel execution of GPUs and TPUs
map well to the paradigm of array programming.
This leads to a gap between available modern hardware architecture and the
necessary tools to leverage their computational power.

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill the resultant gaps led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

Importantly, some projects follow the NumPy API exactly---with Dask,
CuPy, Sparse and JAX the most prominent examples---while others have
NumPy-inspired semantics but, often due to their development history, deviate in places.
Wherever there are such deviations, we often observe users gently requesting,
and developers evolving the software step by step toward the NumPy API.

@@

Ideally, operating on these arrays using NumPy functions or semantics would
simply work, so that end users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays, and so
forth, as appropriate.
To support array operations between external array objects, NumPy
acts as a central coordination mechanism with a well-specified API.
The widespread adoption of the NumPy API is valuable: it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray, by facilitating
the development of specialized solutions and new tools that operate in
concert.

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402
To facilitate \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
However, general adoption has not yet been achieved mainly for compatibility
concerns.
% explain a bit what "compatibility concerns" means ...
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.

>>>>

NumPy provides in-memory, multidimensional, homogeneously typed (i.e.,
single pointer and strided) arrays on CPUs.  It runs on machines
ranging from embedded devices to the world's largest supercomputers,
with performance approaching that of compiled languages.  For the
largest part of its existence, NumPy covered the vast majority of
array computation use cases.

%For algorithms that do not fit the array programming approach,
%or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
%that transparently accelerate bottlenecks.
%These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
%In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure %Python.

@@

In recentl years, there has been a rapid growth in the development of
specialized array processing hardware, including graphics processing
units (GPUs), tensor processing units (TPUs), and field-programmable
gate arrays (FPGAs).  Routinely, data sets are now larger than a
single machine's memory, and may be stored on multiple machines or in
the cloud.  Due to its in-memory data model, NumPy is unable to
utilize such specialized hardware and storage.  However, both
distributed data and the parallel execution of GPUs and TPUs map well
to the \emph{paradigm} of array programming: a gap therefore existed between
available modern hardware architectures and the tools necessary to
leverage their computational power.

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with array is experimental by nature
and, in fact, several promising libraries---such as Caffe, Theano, and
biggus---have already ceased development. And each time that a user
decides to try a new technology or backend library, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

%CuPy, Sparse and JAX the most prominent examples---while others have
%NumPy-inspired semantics but, often due to their development history, deviate in places.
%Wherever there are such deviations, we often observe users gently requesting,
%and developers evolving the software step by step toward the NumPy API.

@@

Ideally, operating on specialized arrays using NumPy functions or semantics would
simply work, so that end users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays, and so
forth, as appropriate.
To support array operations between external array objects, NumPy
therefore added the capability to act as a central coordination
mechanism with a well-specified API.

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402
To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
As an example, consider the arrays shown in ...
\fixme{this part to be fleshed out to match the code in the figure}
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
%However, general adoption has not yet been achieved mainly for compatibility
%concerns.
% explain a bit what "compatibility concerns" means ...
%Thus, we are actively addressing these concerns by revising and extending
%these protocols with the aim of wide interoperability that spans the whole
%scientific Python ecosystem up to and including domain libraries.
